### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@720 fd2da0a4-d968-48df-9cb5-d8c496287cbf

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid (and further subdivisions) of the human brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  Given the ``missing data'' aspect of this
problem, use of the Expectation-Maximization (EM) framework is natural \citep{Dempster1977}.  The
work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended nonparametric
probability model, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

@@

\subsection{Likelihood Implementation}
As mentioned previously in the introduction, different groups have
opted for different likelihood models which have included either
parametric, specifically Gaussian, or nonparametric variations.
However, these approaches are similar in that they require a list
sample of intensity data from the input image(s) and a list of
weighting values for each observation of the list sample from which
the model is constructed.  In general, one may query model
probabilities by passing a given pixel's single intensity (for
univariate segmentation) or multiple intensities (for multivariate
segmentation) to the modeling function, regardless of whether the
function is parameter or non-parametric.  These similarities permitted
the creation of a generic plug-in architecture where classes
describing both parametric and nonparametric observational models are
all derived from an abstract list sample function class.  Three
likelihood classes have been developed, one parametric and one
nonparametric, and are available for usage although one of the
nonparametric classes is still in experimental development.  The
plug-in architecture even permits mixing likelihood models with
different classes during the same run for a hybrid
parametric/non-parametric model although this possibility has yet
to be fully explored.

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\subsubsection{Integration of N4 Bias Correction}
The typical segmentation processing pipeline begins with an intensity
normalization/bias correction step using the recently developed N4
algorithm \citep{Tustison2010} which is based on the popular
nonparametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
several methods proposed in the literature have taken an integrative
view of the segmentation problem by incorporating an intrinsic bias
correction step into the actual workflow.  The
advancements introduced with N4 permit such an adaptive
integration with Atropos.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.  This is done by setting
the `{\ttfamily --weight-image}' option on the N4 command line call
(see Listing \ref{listing:n4}) to the posterior probability image
corresponding to the white matter produced as output in the Atropos
call, i.e. `{\ttfamily Atropos --output}'.  N4 is
currently being added as a standard part of the Insight ToolKit.  The
evaluation section will illustrate inclusion of Atropos, N4 and ANTs
in a brain processing pipeline.  An example of this procedure, using brainweb data with
40\% RF bias, is in figure~\ref{fig:bwebrf40}.  We supply the
information necessary to repeat this example in the script entitled
`{\ttfamily  atroposBwebRF40FigureExample.sh}' which is available in the ANTs toolkit's Atropos
documentation folder as of svn commit 711.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=6in]{Figures/bwebN4ex.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
sequential processing and apply to brainweb single-subject data with
40\% RF bias and 3\% noise.  The $\beta$ for the MRF term is, here,
$0.2$.  Slice 71 of the input data is in (a).  The initial K-means
($K=3$) segmentation is in panel (b).  We use the brain mask to guide
N4 bias correction and produce the image, shown in (c).  We repeat the
K-means segmentation, but with the N4 corrected image as input, and
produce the segmentation in (d).  The average 3-tissue Dice overlap of
result (b) is 0.906 while the average overlap for (d) is 0.954.
Arrows highlight a region of large before-after segmentation
discrepancy.}
\label{fig:bwebrf40}
\end{center}
\end{figure}

>>>>

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid (and further subdivisions) of the human brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  Given the ``missing data'' aspect of this
problem, use of the Expectation-Maximization (EM) framework is natural \citep{Dempster1977}.  The
work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended non-parametric
probability model, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

@@

\subsection{Likelihood Implementation}
As mentioned previously in the introduction, different groups have
opted for different likelihood models which have included either
parametric, specifically Gaussian, or non-parametric variations.
However, these approaches are similar in that they require a list
sample of intensity data from the input image(s) and a list of
weighting values for each observation of the list sample from which
the model is constructed.  In general, one may query model
probabilities by passing a given pixel's single intensity (for
univariate segmentation) or multiple intensities (for multivariate
segmentation) to the modeling function, regardless of whether the
function is parametric or non-parametric.  These similarities permitted
the creation of a generic plug-in architecture where classes
describing both parametric and non-parametric observational models are
all derived from an abstract list sample function class.  Three
likelihood classes have been developed, one parametric and one
non-parametric, and are available for usage although one of the
non-parametric classes is still in experimental development.  The
plug-in architecture even permits mixing likelihood models with
different classes during the same run for a hybrid
parametric/non-parametric model although this possibility has yet
to be fully explored.

@@

OPTIONS: 
     -d, --image-dimensionality 2/3/4
     -i, --input-image inputImageFilename
     -x, --mask-image maskImageFilename
     -w, --weight-image weightImageFilename
     -s, --shrink-factor 1/2/3/4/...
     -c, --convergence [<numberOfIterations=50>,<convergenceThreshold=0.001>]
     -b, --bspline-fitting [splineDistance,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
                           [initialMeshResolution,<splineOrder=3>,<sigmoidAlpha=0.0>,
                           <sigmoidBeta=0.5>]
     -t, --histogram-sharpening [<FWHM=0.15>,<wienerNoise=0.01>,<numberOfHistogramBins=200>]
     -o, --output [correctedImage,<biasField>]
     -h 
     --help 
\end{lstlisting} 
\end{command}
\subsubsection{Integration of N4 Bias Correction}
The typical segmentation processing pipeline begins with an intensity
normalization/bias correction step using the recently developed N4
algorithm \citep{Tustison2010} which is based on the popular
non-parametric nonuniformity intensity normalization (N3) introduced in \cite{Sled1998}.  It was previously mentioned that
several methods proposed in the literature have taken an integrative
view of the segmentation problem by incorporating an intrinsic bias
correction step into the actual workflow.  The
advancements introduced with N4 permit such an adaptive
integration with Atropos.  Recent demonstrations suggest improved
white matter segmentation produces better gain field estimates using
N3 \citep{Boyes2008}.  Thus, when performing 3-tissue segmentation, we
may opt to use, for instance, the posterior probability map of white matter at the current
iteration as a weighted mask for input to N4.  This is done by setting
the `{\ttfamily --weight-image}' option on the N4 command line call
(see Listing \ref{listing:n4}) to the posterior probability image
corresponding to the white matter produced as output in the Atropos
call, i.e. `{\ttfamily Atropos --output}'.  N4 is
currently being added as a standard part of the Insight ToolKit.  The
evaluation section will illustrate inclusion of Atropos, N4 and ANTs
in a brain processing pipeline.  An example of this procedure, using brainweb data with
40\% RF bias, is in figure~\ref{fig:bwebrf40}.  We supply the
information necessary to repeat this example in the script entitled
`{\ttfamily  atroposBwebRF40FigureExample.sh}' which is available in the ANTs toolkit's Atropos
documentation folder as of svn commit 711.
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=6in]{Figures/bwebN4ex.pdf}
\end{tabular}
\caption{\baselineskip 12pt \small We combine N4 and Atropos by simple
sequential processing and apply to brainweb single-subject data with
40\% RF bias and 3\% noise.  The $\beta$ for the MRF term is, here,
$0.2$.  Slice 71 of the input data is in (a).  The initial K-means
($K=3$) segmentation is in panel (b).  We use the brain mask to guide
N4 bias correction and produce the image, shown in (c).  We repeat the
K-means segmentation, but with the N4 corrected image as input, and
produce the segmentation in (d).  The average 3-tissue Dice overlap of
result (b) is 0.906 while the average overlap for (d) is 0.954.
Arrows highlight a region of large before-after segmentation
discrepancy.}
\label{fig:bwebrf40}
\end{center}
\end{figure}
